---
output:
  pdf_document: default
  html_document: default
---
title: 'Worksheet 5: Mapping'
author: 'Lane Riggs'
date: 'March 24, 2023'
---

>I had lots of difficulties with this worksheet. I could never get sf or rgdal to install properly, but I finally got leaflet to. (I was confused during our meeting: Leaflet Providers is NOT leaflet.) I am going to schedule a meeting with you to find out how to install them correctly! I apologize for all of the unanswered questions! I Googled, slacked, and Googled again but could never get the packages to work. 

_This is the fifth in a series of worksheets for History 8510 at Clemson University. The goal of these worksheets is simple: practice, practice, practice. The worksheet introduces concepts and techniques and includes prompts for you to practice in this interactive document. When you are finished, you should change the author name (above), knit your document, and upload it to canvas. Don't forget to commit your changes as you go and push to github when you finish the worksheet._

## Mapping with `ggmap()` and `ggplot2()`

There are many different mapping packages for R. That means there is no single way to create a map in R. Different packages have different strengths and weaknesses and your use of them will depend on your goals and the historical questions you would like to ask. If your project is highly map centric - it may be better suited to ArcGIS which we will not cover in this class. 

```{r message=FALSE, warning=FALSE, include=FALSE}
library(ggplot2) 
library(tidyverse)
library(DigitalMethodsData)
library(tidygeocoder)
```

### Geocoding
The first step in any project is to create geographical data. Depending on the time period you study and the nature of the data, this may or may not be able to be automated. The process of associating geographical coordinates (latitude/longitude) with data is called **geocoding**. There are numerous avenues and services for performing this service. Google Maps and Open Street Maps are the two most common. These services accept an address and return latitude and longitude coordinates. Google Maps does require an API Key which you can sign up for. Typically geocoding with Google costs .5 cents per entry but when you sign up with them, you get $300 in credit per year (at least at the time of writing this - that may change). Although we geocode a very large amount of data with Google on Mapping the Gay Guides, I've never been charged for geocoding. 

However, for the purposes of this class we're going to use Open Street Map's geocoding API because it is open source and therefore free. 

To use the geocoding service, lets first load some data. We'll use the recreation data that we used last week. 

```{r}

rec.data <- read.csv("https://raw.githubusercontent.com/regan008/DigitalMethodsData/main/raw/Recreation-Expenditures.csv")
head(rec.data)

```
Notice in this dataset we have the city state and year but no geographical coordinates if we wanted to map this data. Even if we don't have an exact street address, we can still geocode this data to get coordinates. The function to do that is `geocode()` and we can pass it a city and street. Note the method option, which designates which geocoding service we want to use. 

```{r}

rec.data.coordinates <- rec.data %>% 
  geocode(city = city, state = state, method='osm', lat = latitude, long = longitude)

head(rec.data.coordinates)
```
Now we have latitude and longitude coordinates for our data. 

(@) Use this approach to geocode the `UndergroundRR` data. 

```{r}

data(UndergroundRR) #load dataset

head(undergroundRR) #to see what variables are used for locations 

Underground.Coordinates <- undergroundRR %>% geocode(city = City, state = State, method = 'osm', lat = latitude, long = longitude) #this asks for addresses to be geocoded 

Underground.Coordinates %>%
  select(City, County, State, latitude, longitude)

```

(@) Geocode the Boston Women Voters dataset. Note that the data does include street addresses but because they are broken into parts - street address number, street, etc - you will need to combine them into a full address to pass to the geocoding function. 

```{r}

data("BostonWomenVoters")

BostonWomenVoters.Coordinates <- BostonWomenVoters %>% 
  slice(1:500) %>%
  geocode(state = State.or.Province.of.Birth, street = Street.of.Residence.on.April.1, method= 'osm', lat = latitude, long = longitude)

#this is going to take a very long time to load 
  
#I decided to only look at 500 entries/addresses because otherwise it would take an hour to load
  
BostonWomenVoters.Coordinates %>%
  select(Street.of.Residence.on.April.1, State.or.Province.of.Birth, latitude, longitude)

```

(@) As mentioned above - there are many ways to make a map in R. The `ggmap()` package has a function called `qmplot()` which allows for the quick plotting of maps with data. Look up the documentation for this package and use it to create a plot of the recreational data that we have gecoded above.
```{r}

View(rec.data)

qmplot(longitude, latitude, data = rec.data)

```

### Maps with `ggplot()`

Just like charts in ggplot, maps in ggplot are plotted using a variety of layers. To build a map we need to supply it with geographic data that can use to plot a base map. Your base map will differ depending on the scale of your data, the questions you are asking, and your area of study. For the purposes of this worksheet lets map the gay guides data. Typically you'd need to geocode this data first, but this data has already been geolocated for you. 

First we need to get a basemap. For this example we'll use the `map_data()` function which turns data from the `maps` package into a data frame that is suitable for plotting with ggplot. 

(@) Look at the documentation for `map_data()`. Which geographies does this package provide access to?

> map_data()

>This maps package gives us access to maps of U.S. counties, states, and the continental US; countries like France, Italy, and New Zealand; and a few versions of a world map. The map data package extends the maps package to include datasets from China and Japan, as well. There are two versions of the world map, one is low (mid) resolution, which was imported from the public domain Natural Earth project. Another version of the world map features the Pacific Ocean in the center of the map. 

Lets load the base map data for the US. 
```{r}
usa <- map_data("state")
```

(@) `map_data()` generates a data frame. Take a look at this data frame, what types of data are included? 
 
> This data frame shows the longitude and latitude of locations of more than 40 states. States are in the region column. There is a subregion column, but it doesn't have many values. There are some subregions for Michigan and Virginia. Each region has a group number, so some states are in group 1, group 2, etc. These locations have numerical orders. These numerical orders could relate to the longitude and latitude values for each location. 


We can now pass this data to ggplot to create a simple basemap. When we wanted to create a bar plot using `ggplot()` we called `geom_bar`. When we wanted to create a line chart we used `geom_point()` and `geom_line()`. The sample principle applies here and `ggplot()` provides a geom for maps.
```{r}

ggplot() + 
  geom_map( data = usa, map = usa, aes(long, lat, map_id=region))

```

Now we have a basemap! But what if we want to layer data onto it. Lets add all of the locations in `gayguides` from 1965. First we need to set up our data: 
```{r}

data(gayguides)

gayguides <- gayguides %>% filter(Year == 1965)

```

And then we can use the same mapping code from above but this time we're going to add an additional geom -- `geom_point()` which will point to each of our locations from 1965. 
```{r}

ggplot() + 
  geom_map( data = usa, map = usa, aes(long, lat, map_id=region)) +
  geom_point(data = gayguides, mapping = aes(x=lon, y=lat))

```

(@) This map looks slightly funny, but that is because the data includes entries outside of the contiguous United States. Try filtering out those entries and mapping this again. Can you change the color or size of the points? Can you add a title?

```{r}

gayguides.filter <- gayguides %>%
  filter(state != "AK" & state != "HI") 

#to filter out states that are not in the contiguous U.S. 

ggplot() +
  geom_map(data = usa, map = usa, aes(long, lat, map_id= region)) +
  geom_point(data = gayguides.filter, mapping = aes(x=lon, y=lat, color=type)) +
  labs(title = "1965 Locations")

#it's a little ugly, but it popped up! 

```

(@) Can you map just locations in South Carolina (on a basemap of SC)? 

```{r}

#working on creating dataframes for South Carolina and South Carolina counties

state.sc <- map_data("state")

south.carolina.locations <- subset(state.sc, region == "South Carolina")

counties <- map_data("county")

south.carolina.counties <- subset(counties, region == "South Carolina")


#then filtering gay guides 

gayguides.SC <- gayguides.filter %>%
  filter(state == "SC")

#create a map 

countyborders.sc <- ggplot(data = south.carolina.locations, mapping = aes(x = long, y = lat, group = group)) +
  coord_fixed(1.3) +
  geom_polygon(color = "black", fill = "gray")

map.SC <- gayguides.SC +
  geom_point(data = gayguides.SC, mapping = aes(x=lon, y=lat, color=type, group=city)) +
  geom_text_repel(data = gayguides.SC, mapping = aes(x=lon, y=lat, group=city, label=title)) +
  labs(title = "South Carolina locations")

map.SC
```

(@) Create a map that uses your geocoded data from the Boston Women Voters dataset. 
```{r}

#create dataframes for specific counties/states
states <- map_data("state")

massachusetts <- subset(states, region == "massachusetts")

county <- map_data("county")

ma.counties <- subset(county, region == "massachusetts")

#now we need to create a base map of New Hampshire with borders 

base.map <- ggplot(data = massachusetts, mapping = aes(x = long, y = lat, group = group)) +
  coord_fixed(1.3) +
  geom_polygon(color = "red", fill = "blue") +
  geom_polygon(data = ma.counties, fill = NA, color = "white") +
  geom_polygon(color = "black", fill = NA) 

#Plot these locations on the base map 

massachusetts.map <- base.map +
  geom_point(data = BostonWomenVoters.Coordinates, mapping = aes(x = longitude, y = latitude, color = Country.of.Birth, group = Street.of.Residence.on.April.1)) +
  labs(title = "Boston Women Voters Residence")

massachusetts.map

#for some reason this looks very odd. what did I do wrong here? 
```

Lets return to the recreational data for a minute.

```{r}

head(rec.data)

```
One interesting way to visualize this map might be to plot each location as a point on the map but to use the total_expenditures values to determine the size of the points. 

We can do that by making a small adjustment to the map we made previously. First lets recreate a basic map of all these locations using `ggplot()`
```{r}

ggplot() + 
  geom_map( data = usa, map = usa, aes(long, lat, map_id=region)) +
  geom_point(data = rec.data, mapping = aes(x= longitude, y=latitude))

```

```{r}

ggplot() + 
  geom_map( data = usa, map = usa, aes(x = long, y = lat, map_id=region), fill="white", color="gray") +
  geom_point(data = rec.data, mapping = aes(longitude, latitude, size=total_expenditures))

```

```{r}

library(readr) #you may have to install it using `install.packages()`. 
library(rgdal)
library(sf)
library(maptools)
library(ipumsr)

#NHGIS data is stored in zip files. R has functions for dealing with these but we still need to download the file to our server. Here we're going to write a function that will create a new directory, download the data, and rename it. 
dir.create("data/", showWarnings = FALSE)
get_data <- function(x) {
  download.file("https://github.com/regan008/DigitalMethodsData/blob/main/raw/nhgis0005_shape_simplified.zip?raw=true", "data/nhgis_simplified_shape.zip")
  download.file("https://github.com/regan008/DigitalMethodsData/blob/main/raw/nhgis0005_csv.zip?raw=true", "data/nhgis_data.zip")
}

get_data()

# Change these filepaths to the filepaths of your downloaded extract
nhgis_csv_file <- "data/nhgis_data.zip"
nhgis_shp_file <- "data/nhgis_simplified_shape.zip"

#load the data and shape file into read_nhgis_sf
nhgis <- read_nhgis_sf(
  data_file = nhgis_csv_file,
  shape_file = nhgis_shp_file
)

#filter nhgis so that the map focuses on the 48 contiguous states. 
nhgis <- nhgis_data %>% filter(STATENAM != "Alaska Territory" & STATENAM != "Hawaii Territory")


#plot 
ggplot(data = nhgis, aes(fill = AZF001)) +
  geom_sf() 

```
(@) In the code above, why filter out Hawaii and Alaska? Try uncommenting that line and rerunning the code. What happens? Why might we want to do this? Why might we not want to do this? How does it shape the interpretation?

  > When you filter out Hawaii and Alaska, you only get a map of the mainland United States. We might want to not filter out Hawaii and Alaska so that we see a full map of the United States. But we may not want to filter/unfilter if we want a zoomed in map of the United States. It might be hard for someone to visualize the entire United States if Hawaii and Alaska are not included; so data about the entire United States might not be visually represented in a map.
  
>I think it could be more helpful if there was a way to visualize Hawaii and Alaska, but have the two in a smaller, separate map, and present all of the data in an easy-to-read, accessible color. 
  
  This is a great start. But using AZF001 (Native born males) as the fill does not create a useful visualization. It doesn't give us a sense of the proportion of that data. What would be much better, is if we knew what percentage of the total population foreign born males represented. To get that we have to calculate it. The next few questions will walk build on the code above and walk you through doing this.

(@) First, create a variable called total_male_pop, with the total foreign and native born male population by summing the variables AZF001 and AZF003. 
```{r}

total_male_pop <- nhgis_data %>%
  rowwise() %>%
  mutate(total_male_pop = sum(c(AZF001, AZF003)))
  
```

(@) Now, use the total_male_pop variable and create a variable for the percentage of foreign born males.
```{r}

total_male_pop <- nhgis_data %>%
  rowwise() %>%
  mutate(foreign_males = round(AZF003/total_male_pop), digits = 3)

```

(@) Now map your result. You'll want to replicate the code from the example above, but this time add another layer to the plot - a scale. Here we'll use this scale `scale_fill_continuous("", labels = scales::percent)`

Before you write that code, look up the documentation for the above code (and look at the examples). What does it do? 

> Scale fill continuous is a color scale that ggplot uses where data values can be mapped. With the percentage, we have a variable that goes up to 60%. So, if we were to map this, the colors would be in a gradient from 0 to 60% rather than separate colors.  

Now create the map: 
```{r warning=TRUE}

ggplot(data=nhgis_data, aes(fill = foreign_male)) +
  geom_sf()
  scale_fill_continuous()
```

### Leaflet

In recent years Leaflet has become the most popular open source Javascript library for mapping. In comparison to `ggplot()` the advantage of leaflet is its interactivity. It allows you to zoom in, have pop ups, etc. While `ggplot()` is a powerful tool for static maps and would be useful for a book or journal article, leaflet would make a useful addition to an online digital component.

Like `ggplot()` it works by layering information on top of a basemap. You begin by loading the library and invoking leaflet. 
```{r}
library(leaflet)
my.map <- leaflet()
my.map
```
Now that we've got a leaflet object created, we can add layers to it starting with a basemap. 
```{r}
my.map %>% addTiles()
```
Leaflet refers to tiles - these are sort of like base maps. Next we need to add coordinates. In this example, lets use the coordinates for Dr. Regan's office. 
```{r}
my.map %>% addTiles() %>% addMarkers(lng=-82.836856, lat=34.678286, popup = "Hardin 004")
```

We could also do this with a data frame. Notice that in this example, we use the leaflet function and call the data inside rather than passing the function coordinates manually. We then use the paste function to build out text for a pop up.
```{r}
leaflet(data=rec.data) %>% addTiles() %>% addMarkers(~longitude, ~latitude, popup = paste("The total expenditures in ", rec.data$city, ", ", rec.data$state, " were ",  rec.data$total_expenditures, sep=""))
```


(@) Use leaflet to map a dataset of your choice: 
```{r}

undergroundRR.data <- Underground.Coordinates %>%
  filter(!is.na(Year))

leaflet(data = undergroundRR.data) %>%
  addMarkers(~longitude, ~latitude, popup = paste(undergroundRR.data$City, ", ", undergroundRR.data$State, " in ")) %>%
  addTiles()

#finally got leaflet to work!! Yay!! 

```

(@) Explain what choices you made in building this map? Why might you use leaflet over ggplot? When would ggplot be more desirable? 

> I decided to look at cities and states from the Underground Railroad dataset to see where enslaved peoples escaped using the railroad. I wanted to look at the railroad dataset since I think we look at the gay guides a lot, and I wanted to use a dataset that we haven't worked with as much. 

>I think ggplot would be more desirable if there was a lot of data and you didn't want to pinpoint the data to specific spots. ggplot could also help you if you wanted to visualize your data in other ways, and not just in maps. 


### Exercises
For the next portion of this worksheet I'd like you to look back at the email about the National Parks Data. Using this data (link below) you should use ggplot (charts, maps) and other mapping tools to come up with several ways to visualize it based on the points outlined in Dr. Barczewski's email. You should try to think about this from the perspective of her larger book project, how could you visualize this data to help her make a compelling argument? 

```{r}
parks <- read.csv("https://raw.githubusercontent.com/regan008/DigitalMethodsData/main/raw/parks-geocoded.csv")
```

```{r}

parks.regions <- parks %>%
  mutate(region = case_when(country %in% parks ~ "Europe", 
                            country == "United Kingdom" ~ "United Kingdom", 
                            country == "Canada" ~ "Canada",
                            country == "South Africa" ~ "South Africa"))

parks.regions %>%
  group_by(region, year, geoAddress) %>%
  ggplot(aes(x = year, y = geoAddress, color = region, size = total_sq_kilometers)) +
  geom_point() +
  ggtitle("Park Size")

#I could not get this to work perhaps because of dev.controls. Again, this may be an issue related to rgdal or sf. I'm sorry I couldn't do as much as others! 


```
